### 2021-04-13 introduction.tex
Rewrote intro.

\begin{document}
\section{Introduction}
Session types have been studied for many programming paradigms, including concurrent and functional paradigms. In particular, \emph{binary} session types have been defined for the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a process calculus for communication and concurrency; and later on for Good Variation~\cite[GV]{wadler15,lindleymorris15}, a session-typed concurrent $\lambda$-calculus.
Binary session types guarantee communication safety and session fidelity, namely that communication proceeds without any type mismatch and it follows the predefined protocol.
Another key property in a concurrent setting is \emph{deadlock freedom}, stating that agents communicate without getting stuck in cyclic dependencies. Deadlock freedom is a more involving property and binary session types \emph{alone} are not enough to rule out deadlocks in cyclic structures of different interleaved sessions. Several techniques have been developed to address deadlock freedom in session-typed $\pi$-calculus and concurrent $\lambda$-calculus.

With regards to $\pi$-calculus, a growing line of work leverages the Curry-Howard correspondences between session-typed $\pi$-calculus and intuitionistic~\cite{cairespfenning10} or classical linear logic~\cite{wadler14}, which guarantees deadlock freedom by restricting processes to only trees. The tree structure is achieved by merging constructs for parallel composition and channel restriction under the logical cut rule, which forces processes to share \emph{only one} channel for communication.
Recent developments in the CP~\cite{wadler14} line of work have led to various approaches to decoupling these constructs. One way is by maintaining a strong correspondence to classical linear logic and allowing only tree-structured processes, such as in Hypersequent CP~\cite[HCP]{kokkemontesi19popl,kokkemontesi19tlla}, Linear Compositional Choreographies~\cite[LCC]{CarboneMS18} or Classical Transitions~\cite[CT]{MP18}. Another way is by weakening the correspondence to classical linear logic in exchange for a more expressive language which allows cyclic-structured processes, such as in Priority CP~\cite[PCP]{dardhagay18}. PCP decouples CP's cut rule into two separate constructs: a construct for parallel composition, by introducing a mix rule; and a construct for restriction, by introducing a cycle rule. In order to restore deadlock freedom, PCP uses priorities following Kobayashi~\cite{kobayashi06} and Padovani~\cite{padovani14}. Priorities establish an \emph{order} of actions and are used to rule out bad interleavings of channels. By leveraging priorities, Dardha and Gay prove cycle-elimination of PCP (following the standard cut-elimination proof for classical linear logic). Deadlock freedom then follows as a corollary of the cycle-elimination result.

With regards to GV, the original work on session types in GV~\cite{gayvasconcelos12} did not satisfy deadlock freedom. Later on, calculi in the GV family~\cite{wadler15,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, \ie, by combining channel creation and thread spawning into a single operation, called fork. The fork construct is the term representation of the cut construct in CP, which as discussed above, evaluates to a channel restriction combined with a parallel composition. As with CP, this is restrictive because it limits process structures to only trees.

We aim to further investigate GV and deadlock freedom, as there are several benefits of working with a functional language as opposed to name-passing process calculi. Functional languages support higher-order functions, and have a capability for abstraction not usually present in process calculi. Working within a functional language allows to derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \ie, deriving internal/external choice via sum types. In concurrent functional calculi, there is a clear separation between a program that the user writes, and a configuration, which is the state of a system as it evaluates. In process calculi, these are conflated. Finally, functional calculi bring us a step closer to transferring our work to real-world (functional) programming languages~\cite{KD21b}.
Moreover, the benefit of working with GV as opposed to other session-typed functional languages is that GV can be translated into CP~\cite{wadler14}, thus guaranteeing the same formal properties as CP, \eg, deadlock freedom, albeit only for tree-structured processes.
Following the above observations, we thus pose our research question for GV.%

\textbf{RQ:} \emph{Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?}

We address this research question by following the line of work from CP to Priority CP, and present
Priority GV (PGV), a~variant of GV which decouples channel creation from thread spawning, thus allowing cyclic-structured processes and restores deadlock freedom via priorities. This closes the circle of the connection between CP and GV~\cite{wadler14}, and their priority-based versions, PCP~\cite{dardhagay18} and PGV.%
In this paper we make the following main contributions:
\begin{enumerate}
	\item \textbf{Priority GV}. We present Priority GV (PGV), a session-typed functional language with priorities (\cref{sec:pgv}) and prove subject reduction (\cref{thm:pgv-subject-reduction-confs}) and progress (\cref{thm:pgv-closed-progress-confs}).
	PGV addresses several problems in the original GV language, most notably:
 	\begin{enumerate*}
	\item we do not require the pseudo-type $S^\sharp$; and
	\item our structural congruence is type preserving.
	\end{enumerate*}
	PGV answers our research question as it allows cyclic-structured binary session-typed processes that are deadlock free.
	\item \textbf{Updated PCP}. We present an updated version of Priority CP~\cite{dardhagay18} in \cref{sec:pcp}. We remove commuting conversions and move away from reduction as cut elimination, and present a reduction relation based on structural congruence, as it is standard in process calculi.
	\item \textbf{Translation from PCP to PGV}. We present a \emph{sound and complete encoding} of PCP in PGV in~\cref{sec:pcp}, where we prove the encoding preserves typing (\cref{thm:pcp-to-pgv-confs-preservation}) and satisfies operational correspondence (\cref{thm:pcp-to-pgv-operational-correspondence-soundness,thm:pcp-to-pgv-operational-correspondence-completeness}).
\end{enumerate}
\end{document}

>>>>

\begin{document}
\section{Introduction}
Session types~\cite{honda93,takeuchihonda94,hondavasconcelos98} are types for protocols. Regular types ensure functions are used according to their specification. Session types ensure \emph{communication channels} are used according to their protocols. Session types have been studied in many settings. For instance, in the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a foundational calculus for communication and concurrency, and in concurrent $\lambda$-calculi~\cite{gayvasconcelos10}, including the focus of our paper: Good Variation~\cite[GV]{wadler15,lindleymorris15}.

GV is a concurrent $\lambda$-calculus with \emph{binary} session types, where each channel is shared between exactly two processes. Binary session types guarantee two crucial properties \emph{communication safety}---\eg, if the protocol says to transmit an integer, you transmit an integer---and \emph{session fidelity}---\eg, if the protocol says send, you send. A third crucial property is \emph{deadlock freedom}, which ensures that processes do not have cyclic dependencies---\eg, when two processes wait for each other to send a value. Binary session types \emph{alone} are insufficient to rule out deadlocks arising from interleaved sessions, but several additional techniques have been developed to guarantee deadlock freedom in session-typed $\pi$-calculus and concurrent $\lambda$-calculus.

In the $\pi$-calculus literature, a growing line of work leverages the Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition---lack of modularity, difficulty with typing structural congruence or with label-transition semantics, \etc---which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad interleavings. Dardha and Gay~\cite{dardhagay18} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

CP is related to GV via a pair of translations which satisfy simulation~\cite{lindleymorris16}, and which can be tweaked to satisfy reflection. Consequently, the two calculi share the same strong guarantees. GV achieves deadlock freedom via a similar syntactic restriction: it combines channel creation and thread spawning into a single operation, called ``fork'', which closely corresponds to the cut construct in CP. Unfortunately, as with CP, this syntactic restriction has many downsides.

Our aims are to develop the expressiveness of GV but maintain deadlock freedom. We chose GV for several reasons, some of which apply more generally as benefits to working within a concurrent $\lambda$-calculus as opposed to a name-passing process calculus. Concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. \emph{Case in point}: we were able to embed the theory developed in this paper in Linear Haskell~\cite{bernardyboespflug18} without much effort~\cite{KD21b}. For us, these advantages outweigh the benefits of process calculi, \eg, their succinctness.

The benefits of working with GV specifically, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its strong formal properties. We thus pose our research question for GV:
\begin{quotation}
\textbf{RQ:} \emph{Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?}
\end{quotation}

We follow the line of work from CP to Priority CP, and present Priority GV (PGV), a~variant of GV which decouples channel creation from thread spawning, thus allowing cyclic-structured processes, but which nonetheless guarantees deadlock freedom via priorities. This closes the circle of the connection between CP and GV~\cite{wadler14}, and their priority-based versions, PCP~\cite{dardhagay18} and PGV.
We make the following main contributions:
\begin{enumerate}
\item \textbf{Priority GV}. We present Priority GV (\secref{sec:pgv}, PGV), a session-typed functional language with priorities, and prove subject reduction (\cref{thm:pgv-subject-reduction-confs}) and progress (\cref{thm:pgv-closed-progress-confs}). We addresses several problems in the original GV language, most notably:
  \begin{enumerate*}
  \item PGV does not require the pseudo-type $S^\sharp$; and
  \item its structural congruence is type preserving.
  \end{enumerate*}
  PGV answers our research question positively as it allows cyclic-structured binary session-typed processes that are deadlock free.
\item \textbf{Updated PCP}. We present an updated version of Priority CP~\cite{dardhagay18} (\secref{sec:pcp}). We remove commuting conversions and move away from reduction as cut elimination, and present a reduction relation based on structural congruence, as it is standard in process calculi.
\item \textbf{Translation from PCP to PGV}. We present a \emph{sound and complete encoding} of PCP in PGV (\secref{sec:pcp}). We prove the encoding preserves typing (\cref{thm:pcp-to-pgv-confs-preservation}) and satisfies operational correspondence (\cref{thm:pcp-to-pgv-operational-correspondence-soundness,thm:pcp-to-pgv-operational-correspondence-completeness}).
\end{enumerate}
\end{document}
